A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
More
Updates from Overleaf

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analysis widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
if we like this].
We hope that the ARG provided here will similarly
enable a range recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements fully integrating
recombination into pathogen genomics.

The second primary contribution of this work is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
Similarly, detecting and fully incorporating recombination 
in real-time at pandemic scale extends the 
demonstrated utility of a comprehensive
genealogy~\cite{Hinrichs2024,McBroome2024}, clarifying the 
role of recombination and enabling timely analysis.
Given the novelty of sc2ts and the inferred ARG, we have focused 
significant attention here on methodological issues. 
Through these detailed analyses we validate the ARG itself,
ensuring that it is a trustworthy basis for future research,
and deepen our understanding of the inference method, and 
how it may be adapted and improved (see Limitations of the Study).

The secondary contributions of this work are the initial 
insights into the evolution of SARS-CoV-2 obtained by studying
the sc2ts ARG. Firstly, we have clarified the relationships among
the recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Secondly, we have used the ARG to study breakpoint intervals 
for recombination events, adding a novel temporal dimension
and demonstrating a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage heterozygosity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we have provided a stringent lower
bound on the amount of recombination that occured during
the pandemic phase of SARS-CoV-2.
Together, these  underscore the utility of the ARG itself, and suggest
several important directions for future research.

@@

@@

Jackson et al. described
four groups of recombinants (named A to D) and four singleton recombinants.
Of the 16 samples analyzed by Jackson et al., 12 were included in the ARG;
the remaining 4 were excluded
because the sequences did not pass the QC filtering criteria or
because of a high HMM cost.
Groups A-D are summarized in Table~\ref{tab:jackson} and
show excellent concordance
in terms of the parental Pango lineages and breakpoint intervals.
All four groups are directly associated with
strongly supported recombination events in the ARG
in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
In the ARG, the singleton CAMC-CBA018 has
the parents B.1.177+B.1.1.7 and
the breakpoint interval 16,177--20,133, and
is strongly supported by 14 averted mutations.
In the Jackson et al. results, this singleton is
almost identical with the same parent lineages and
a breakpoint interval of 11,395–-21,993.
In the ARG, the singleton MILK-103C712 has
the parents B.1.177.16+B.1.1.7 and
a breakpoint interval of 24915--27972,
again strongly supported by 11 averted mutations.
In the Jackson et al. results,
this lineage has the same parents,
with the breakpoint interval 26,800–-27,974.
In the ARG, the singleton QEUH-1067DEF has
the parents B.1.1.7+B.1.177.9 and
a breakpoint interval of 7,729--10,870,
supported by 11 averted mutations.
In the Jackson et al.\ results,
this lineage has the same parents,
with the breakpoint interval 6,953--10,872.

@@

@@

\newcommand{\cprheight}{4.0em}

@@

\renewcommand{\cprheight}{1.5em}
% Add some more space between rows. Note this it would be sligntly nicer if the 
% label was aligned vertically in the center of the row, but this is good
% enough for now.
\renewcommand{\arraystretch}{2}

>>>>

\section*{DISCUSSION}
There are two major contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analysis widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
if we like this].
We hope that the ARG provided here will similarly
enable a range recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements fully integrating
recombination into pathogen genomics.

The second major contribution is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
By capturing recombination in real time at a global scale, our approach extends the established utility of comprehensive genealogies~\cite{Hinrichs2024,McBroome2024}, offers clear insight into recombination dynamics, and supports timely investigation of .
The novelty of sc2ts and the inferred ARG motivates our  focus on methodological details. 
Through these detailed analyses we validate the ARG itself,
ensuring that it is a trustworthy basis for future research,
and deepen our understanding of the inference method, and 
how it may be adapted and improved (see Limitations of the Study).

Study of the sc2ts ARG provides insights in the  contributions of this work are provided by our 
insights into the evolution of SARS-CoV-2 obtained by studying
the sc2ts ARG. Firstly, we have clarified the relationships among
the recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Secondly, we have used the ARG to study breakpoint intervals 
for recombination events, adding a novel temporal dimension
and demonstrating a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage heterozygosity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we have provided a stringent lower
bound on the amount of recombination that occured during
the pandemic phase of SARS-CoV-2.
Together, these  underscore the utility of the ARG itself, and suggest
several important directions for future research.

@@

Jackson et al. described
four groups of recombinants (named A to D) and four singleton recombinants.
Of the 16 samples analyzed by Jackson et al., 12 were included in the ARG;
the remaining 4 were excluded
because the sequences did not pass the QC filtering criteria or
because of a high HMM cost.
Groups A-D are summarized in Table~\ref{tab:jackson} and
show excellent concordance
in terms of the parental Pango lineages and breakpoint intervals.
All four groups are directly associated with
strongly supported recombination events in the ARG
in terms of the number of averted mutations (A: 16, B: 6, C: 30, D: 18).
In the ARG, the singleton CAMC-CBA018 has
the parents B.1.177+B.1.1.7 and
the breakpoint interval 16,177--20,133, and
is strongly supported by 14 averted mutations.
In the Jackson et al. results, this singleton is
almost identical with the same parent lineages and
a breakpoint interval of 11,395–-21,993.
In the ARG, the singleton MILK-103C712 has
the parents B.1.177.16+B.1.1.7 and
a breakpoint interval of 24915--27972,
again strongly supported by 11 averted mutations.
In the Jackson et al. results,
this lineage has the same parents,
with the breakpoint interval 26,800--27,974.
In the ARG, the singleton QEUH-1067DEF has
the parents B.1.1.7+B.1.177.9 and
a breakpoint interval of 7,729--10,870,
supported by 11 averted mutations.
In the Jackson et al.\ results,
this lineage has the same parents,
with the breakpoint interval 6,953--10,872.

@@

@@

\newcommand{\cprheight}{1.5em}
% Add some more space between rows. Note this it would be sligntly nicer if the 
% label was aligned vertically in the center of the row, but this is good
% enough for now.
\renewcommand{\arraystretch}{2}

@@

\renewcommand{\cprheight}{4.0em}

@@
